Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2011-06-30 conclusions.tex
Adjusted spacing between some figures and their units
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1231 ccdd05d5-56c1-461a-bf98-85f6876523b4

The \lloid{} method is as fast as standard \fft{} convolutions but allows for
nearly latency-free, real-time operation.  We anticipate requiring $\sim$~100
modern multi-core computers to search for binary neutron stars using
coincident \GW\ data from a multi-detector network.  In the future, additional
computational savings could be acheived by conditionally reconstructing the
\SNR{} time-series only during times when a composite detection statistic
crosses a threshold~\citep{svd-compdetstat}.  However, the anticipated required
number of computers is well within the current computing capabilities of the
\LIGO{} Data
Grid\footnote{\url{https://www.lsc-group.phys.uwm.edu/lscdatagrid/}}.

\CBC\ events may be the progenitors of some short hard \GRB{}s and are expected
to be accompanied by prompt \EM{} signals.  Rapid alerts to the broader
astronomical community will improve the chances of detecting an \EM{}
counterpart in bands from gamma-rays down to radio.  It will be extremely
challenging for all but the rarest events to localize the source well enough in
advance to detect the onset of prompt emission, however localization prior to
the onset of extended emission on the 10--100\,s time-scale is more easily
achieved and will benefit from the full bandwidth and \SNR\ of the \GW{} event
at merger.  Probing even the extended emission timescale requires very low
latency filtering schemes such as the one described here.

>>>>

The \lloid{} method is as fast as standard \fft{} convolutions but allows for
nearly latency-free, real-time operation.  We anticipate requiring $\sim$100
modern multi-core computers to search for binary neutron stars using
coincident \GW\ data from a multi-detector network.  In the future, additional
computational savings could be acheived by conditionally reconstructing the
\SNR{} time-series only during times when a composite detection statistic
crosses a threshold~\citep{svd-compdetstat}.  However, the anticipated required
number of computers is well within the current computing capabilities of the
\LIGO{} Data
Grid\footnote{\url{https://www.lsc-group.phys.uwm.edu/lscdatagrid/}}.

\CBC\ events may be the progenitors of some short hard \GRB{}s and are expected
to be accompanied by prompt \EM{} signals.  Rapid alerts to the broader
astronomical community will improve the chances of detecting an \EM{}
counterpart in bands from gamma-rays down to radio.  It will be extremely
challenging for all but the rarest events to localize the source well enough in
advance to detect the onset of prompt emission, however localization prior to
the onset of extended emission on the 10--100~s time-scale is more easily
achieved and will benefit from the full bandwidth and \SNR\ of the \GW{} event
at merger.  Probing even the extended emission timescale requires very low
latency filtering schemes such as the one described here.

### 2011-06-30 method.tex
Adjusted spacing between some figures and their units
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1231 ccdd05d5-56c1-461a-bf98-85f6876523b4

For concreteness and simplicity, consider an inspiral waveform in the
quadrapole approximation, for which the time-frequency relation~\citep{kidder1992, findchirppaper} is
%
\begin{equation} \label{eq:fgw}
%
f(t) = \frac{1}{\mathcal{\pi M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5 \upmu\mathrm{s}$) and $t$ is the time relative to the
coalescence of the binary.
This monotonic time-frequency relationship allows us
to choose time slice boundaries that require substantially less bandwidth at
early times in the inspiral.

>>>>

For concreteness and simplicity, consider an inspiral waveform in the
quadrapole approximation, for which the time-frequency relation~\citep{kidder1992, findchirppaper} is
%
\begin{equation} \label{eq:fgw}
%
f(t) = \frac{1}{\pi \mathcal{M}} \left[ \frac{5}{256}\frac{\mathcal{M}}{t}
\right]^{3/8}.
%
\end{equation}
%
Here, $\mathcal{M}$ is the chirp mass of the binary in units of time (where $G
M_\odot / c^3 \approx 5$~$\upmu$s) and $t$ is the time relative to the
coalescence of the binary.
This monotonic time-frequency relationship allows us
to choose time slice boundaries that require substantially less bandwidth at
early times in the inspiral.

### 2011-06-30 prospects.tex
Adjusted spacing between some figures and their units
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1231 ccdd05d5-56c1-461a-bf98-85f6876523b4

\begin{figure}[h]
\plotone{figures/snr_in_time}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that could be detectable by Advanced \LIGO\ a given number of seconds
before coalescence.  The heavy solid line is the most realistic yearly rate
estimate.  The shaded region represents the 5 to 95\% confidence interval
arising from uncertainty in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}
%
Examining the signal to noise ratio (\SNR{}) stream for threshold crossings
before the \GW{} signal leaves the detection band makes it possible to trade
some \SNR{} and sky localization accuracy for pre-merger detection.
Figure~\ref{fig:earlywarning} shows projected early detectability rates for
\textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40\,events\,yr$^{-1}$. $\sim$10\,yr$^{-1}$ will be detectable
within 10\,s of merger and $\sim$5\,yr$^{-1}$ will be detectable within 25\,s of
merger if analysis can proceed with near zero latency. These rates are shown
with their substantial uncertainties in Figure~\ref{fig:earlywarning}.

@@

\begin{table}[h]
\caption{\label{table:sky-localization-accuracy}Horizon distance, \SNR\ at
merger, and area of 90\% confidence at selected times before merger for sources
with expected detectability rates of 40, 10, 1, and 0.1\,yr$^{-1}$.}
\begin{center}
\begin{tabular}{rrrrrrr}
\tableline\tableline
rate & horiz. & final & \multicolumn{4}{c}{$A$(90\%) (deg$^2$)} \\
\cline{4-7}
yr$^{-1}$ & (Mpc) & \SNR\ & 25 s & 10 s & 1 s & 0 s \\
\tableline
40\phd\phn & 445 & 8.0 & 15500 & 3100 & 200 & 9.6 \\
10\phd\phn & 280 & 12.7 & 6100 & 1200 & 78 & 3.8 \\
1\phd\phn & 130 & 27.4 & 1300 & 260 & 17 & 0.8 \\
0.1 & 60 & 58.9 & 280 & 56 & 3.6 & 0.2 \\
\tableline
\end{tabular}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{figures/loc_in_time}
\caption{\label{fig:sky-localization-accuracy}Area of the 90\% confidence
region as a function of time before coalescence for sources with anticipated
detectability rates of 40, 10, 1, and 0.1\,yr$^{-1}$. The heavy dot indicates
the time at which the accumulated \SNR\ exceeds a threshold of 8.}
\end{figure}

\EM\ followup requires estimating the location of the \GW{} source. The localization
uncertainty can be estimated from the uncertainty in the time of arrival of the \GW{}s,
which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}.  Table~\ref{fig:earlywarning} and
Figure~\ref{fig:sky-localization-accuracy} show the estimated minimum 90\%
confidence area versus time of the loudest coalescence events detectable by
Advanced \LIGO{} and Advanced Virgo.  Once per year, we expect to observe an
event with a final \SNR\ of $\approx 27$ whose location can be constrained to about
1300\,deg$^2$ (3.1\% of the sky) within 25\,s of merger,
260\,deg$^2$ (0.63\% of the sky) within 10\,s of merger, and
0.82\,deg$^2$ (0.0020\% of the sky) at merger.

It is infeasible to search hundreds of square degrees for a prompt counterpart.
For reference, \textsc{lsst} should see first light around the same time as
Advanced \LIGO{} and will have an unparalleled field of view of 9.6 deg$^2$
\citep{2008arXiv0805.2366I}.  However, 
it is possible to reduce the localization uncertainty by only looking at
galaxies from a catalog that lie near the sky location and luminosity distance
estimate from the \GW{} signal~\citep{galaxy-catalog} as was done in S6/VSR3.
Within the expected Advanced \LIGO{} \textsc{ns}--\textsc{ns} horizon distance,
the number of galaxies that can produce a given signal amplitude is much larger
than in Initial \LIGO{} and thus the catalog will not be as useful
for downselecting pointings generally. However, exceptional \GW{} sources will
necessarily be extremely nearby. Within this reduced volume there will be fewer
galaxies to consider for a given candidate and catalog completeness will be
less of a concern.  This should reduce the 90\% confidence area substantially.
Also future \GW{} detectors, should they be built, will improve the \SNR\ and
likewise the error box.

>>>>

\begin{figure}[h]
\plotone{figures/snr_in_time}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that could be detectable by Advanced \LIGO\ a given number of seconds
before coalescence.  The heavy solid line is the most realistic yearly rate
estimate.  The shaded region represents the 5 to 95\% confidence interval
arising from uncertainty in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}
%
Examining the signal to noise ratio (\SNR{}) stream for threshold crossings
before the \GW{} signal leaves the detection band makes it possible to trade
some \SNR{} and sky localization accuracy for pre-merger detection.
Figure~\ref{fig:earlywarning} shows projected early detectability rates for
\textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$. $\sim$10~yr$^{-1}$ will be detectable
within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable within 25~s of
merger if analysis can proceed with near zero latency. These rates are shown
with their substantial uncertainties in Figure~\ref{fig:earlywarning}.

@@

\begin{table}[h]
\caption{\label{table:sky-localization-accuracy}Horizon distance, \SNR\ at
merger, and area of 90\% confidence at selected times before merger for sources
with expected detectability rates of 40, 10, 1, and 0.1~yr$^{-1}$.}
\begin{center}
\begin{tabular}{rrrrrrr}
\tableline\tableline
rate & horiz. & final & \multicolumn{4}{c}{$A$(90\%) (deg$^2$)} \\
\cline{4-7}
yr$^{-1}$ & (Mpc) & \SNR\ & 25 s & 10 s & 1 s & 0 s \\
\tableline
40\phd\phn & 445 & 8.0 & 15500 & 3100 & 200 & 9.6 \\
10\phd\phn & 280 & 12.7 & 6100 & 1200 & 78 & 3.8 \\
1\phd\phn & 130 & 27.4 & 1300 & 260 & 17 & 0.8 \\
0.1 & 60 & 58.9 & 280 & 56 & 3.6 & 0.2 \\
\tableline
\end{tabular}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{figures/loc_in_time}
\caption{\label{fig:sky-localization-accuracy}Area of the 90\% confidence
region as a function of time before coalescence for sources with anticipated
detectability rates of 40, 10, 1, and 0.1~yr$^{-1}$. The heavy dot indicates
the time at which the accumulated \SNR\ exceeds a threshold of~8.}
\end{figure}

\EM\ followup requires estimating the location of the \GW{} source. The localization
uncertainty can be estimated from the uncertainty in the time of arrival of the \GW{}s,
which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}.  Table~\ref{fig:earlywarning} and
Figure~\ref{fig:sky-localization-accuracy} show the estimated minimum 90\%
confidence area versus time of the loudest coalescence events detectable by
Advanced \LIGO{} and Advanced Virgo.  Once per year, we expect to observe an
event with a final \SNR\ of $\approx$27 whose location can be constrained to about
1300~deg$^2$ (3.1\% of the sky) within 25~s of merger,
260~deg$^2$ (0.63\% of the sky) within 10~s of merger, and
0.82~deg$^2$ (0.0020\% of the sky) at merger.

It is infeasible to search hundreds of square degrees for a prompt counterpart.
For reference, \textsc{lsst} should see first light around the same time as
Advanced \LIGO{} and will have an unparalleled field of view of 9.6~deg$^2$
\citep{2008arXiv0805.2366I}.  However, 
it is possible to reduce the localization uncertainty by only looking at
galaxies from a catalog that lie near the sky location and luminosity distance
estimate from the \GW{} signal~\citep{galaxy-catalog} as was done in S6/VSR3.
Within the expected Advanced \LIGO{} \textsc{ns}--\textsc{ns} horizon distance,
the number of galaxies that can produce a given signal amplitude is much larger
than in Initial \LIGO{} and thus the catalog will not be as useful
for downselecting pointings generally. However, exceptional \GW{} sources will
necessarily be extremely nearby. Within this reduced volume there will be fewer
galaxies to consider for a given candidate and catalog completeness will be
less of a concern.  This should reduce the 90\% confidence area substantially.
Also future \GW{} detectors, should they be built, will improve the \SNR\ and
likewise the error box.
